---
title: "Opioid Overdose Report"
output: word_document
---

```{r echo=FALSE}
filtered <-
      data %>%
      filter(
        Indicator.Number == input$indicatorInput,
        variable == input$yearInput
      )
    
    newslope <- filter(slopes, Indicator.Number == input$indicatorInput)
    
    #merge state shape data with indicator data
    graphdata <- inner_join(filtered, states, by = "Awardee")
```
###`r filtered[1,"Indicator.Name"]` in `r input$yearInput`

The map below shows reported levels of `r tolower(filtered[1,"Indicator.Name"])` in `r input$yearInput` for all PfS / DDPI states. The color of each state represents the relative number of `r tolower(filtered[1,"Indicator.Name"])` reported in `r input$yearInput` for each state.

The circles on each state represent the annual rate of change in `r tolower(filtered[1,"Indicator.Name"])` observed between 2013 and 2015. Green circles indicate states that are improving, and red circles indicate states that are getting worse. Large circles indicate a higher rate of change.
```{r fig.width=12, fig.height=6, echo=FALSE}

    ggplot(data = graphdata) + 
      geom_polygon(data = map_data("state"), aes(x=long, y = lat, fill = value, group = group), fill = "grey", color = "white") +
      geom_polygon(aes(x = long, y = lat, fill = value, group = group), color = "grey40") + 
      scale_fill_gradient(low = 'lightblue', high = 'darkblue', name = paste(input$yearInput[1],"Value"), 
                           limits=c(0, max(subset(slopes, Indicator.Number == input$indicatorInput)$value))) +
      geom_point(data = newslope, aes(long, lat, size = (abs(Slope)), color = sign, shape = sign), fill = "white") +
      scale_size(name = "Yearly Change") +
      scale_shape_manual(values=c(19, 19), name = "Trend", labels = c("Getting Better","Getting Worse")) +
      scale_color_manual(values=c("green", "red"), name = "Trend", labels = c("Getting Better","Getting Worse")) +
      coord_fixed(1.3) +
      theme(axis.line=element_blank(),
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            axis.title.x=element_blank(),
            axis.title.y=element_blank(),
            panel.background=element_blank(),
            panel.border=element_blank(),
            panel.grid.major=element_blank(),
            panel.grid.minor=element_blank(),
            plot.background=element_blank())
```

*Note*: The data presented here are required indicators provided by PfS/DDPI states. They represent age-adjusted rates per 100,000 residents from the corresponding year.

Overall, the average rate of `r tolower(filtered[1,"Indicator.Name"])` across all PfS / DDPI states was `r round(mean(filtered$value, na.rm = TRUE), digits = 1)` per 100,000 residents in `r input$yearInput`.



\newpage
```{r echo=FALSE,warning=FALSE,message=FALSE}
oldvalue <- mean(filtered$value, na.rm = TRUE)

filtered <-
      data %>%
      filter(
        Indicator.Number == input$indicatorInput
      )

regression <- lm(value ~ variable, data = filtered)
slopesign <- ifelse(regression$coeff[2] >= 0, "increasing", "decreasing")
newvalue <- regression$coeff[1] + (regression$coeff[2]*2018)
library(scales)
```
####Nationwide forecast for `r tolower(filtered[1,"Indicator.Name"])`

Nationwide, the number of `r tolower(filtered[1,"Indicator.Name"])` has been `r slopesign` by a rate of about `r round(regression$coeff[2], 1)` per 100,000 annually, on average.

If this trend continues, by 2018 the average level of `r tolower(filtered[1,"Indicator.Name"])` will be `r round(newvalue,1)` per 100,000 residents - an estimated `r format(round(newvalue*3224,0),scientific=FALSE)` people annually. This represents a `r percent((newvalue-oldvalue)/oldvalue)` change since `r input$yearInput`.

The graph below displays the predicted forecast for `r tolower(filtered[1,"Indicator.Name"])` based on PfS / DDPI data from 2013 to 2015.

```{r fig.width=12, fig.height=3, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(filtered,aes(variable,value)) +
      stat_summary(fun.data = "mean_se", color = "blue", size = 1) + 
      geom_smooth(method='lm',fullrange=TRUE) + 
      xlim(2013, 2018) +
      labs(title = paste("Nationwide trend for",tolower(filtered[1,"Indicator.Name"])), x = "Year", y = paste("National Average"))

```

####State comparisons for `r tolower(filtered[1,"Indicator.Name"])`

Across all years surveyed, the state with the lowest level of `r tolower(filtered[1,"Indicator.Name"])` was `r filtered[which.min(filtered$value),"Awardee"]` in `r filtered[which.min(filtered$value),"variable"]` with a rate of `r round(filtered[which.min(filtered$value),"value"],1)`.

The state with the highest level of `r tolower(filtered[1,"Indicator.Name"])` was `r filtered[which.max(filtered$value),"Awardee"]` in `r filtered[which.max(filtered$value),"variable"]` with a rate of `r round(filtered[which.max(filtered$value),"value"],1)`.

Overall, `r newslope[which.max(newslope$Slope),"Awardee"]` had the highest rate of change with an average annual change of `r round(newslope[which.max(newslope$Slope),"Slope"],1)`.

\newpage
A table of the data used to generate this report is included below.

``` {r echo=FALSE, warning=FALSE}
    short <- filter(short, Indicator.Number == input$indicatorInput)
    short["Merge"] <- mutate_all(short["Awardee"], funs(tolower))
    #change state names to lower-case for merge
    slopes <- filter(slopes, Indicator.Number == input$indicatorInput, variable == 0)
    colnames(slopes)[3] <- "Merge"
    display <- left_join(short,slopes,"Merge")
    myvars <- c("Awardee","2013","2014","2015","Slope")
    display <- display[myvars]
    display <- arrange(display,Awardee)
    library(knitr)
    kable(display, digits = 2)
```